---
title: "Balance responses in MS"
author:
- J. Lucas McKay
- Daniel S. Peterson
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
bibliography: bibliography.bib
csl: nature.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
# set path variables
source("../paths.m")

# if csv files do not exist, run matlab batch steps
if (!file.exists("../csv/c.csv")|!file.exists("../csv/m.csv")){
  system(paste(matlabCommand, "'cd ../m; loadData; exit'"))
}
```

# Introduction
This document describes the analysis of standing balance responses in people with and without multiple sclerosis (MS) using wavelet-based functional ANOVA (wfANOVA[@mckay2013]).
MS has profound effects on balance,[@vanliew2019] including altering protective stepping behaviors.[@peterson2016]
The main objective of this study was to examine the impact of MS on the pattern of magnitude and timing of muscle activation during feet-in-place balance responses.

# Methods

## Data sources

This study performed a secondary analysis on existing electromyographic recordings during balance responses in people with and without MS.
Clinical and demographic characteristics of the study sample are summarized in Table 1.

```{r, message=FALSE, warning=FALSE}
library(vroom)
library(dplyr)
library(stringr)
library(lubridate)
`%contains%` = function(s, pattern) grepl(pattern, s)
`%omits%` = function(s, pattern) !grepl(pattern, s)

# load participant-level information
m = vroom("../csv/m.csv") %>% mutate(Group = "MS") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)
c = vroom("../csv/c.csv") %>% mutate(Group = "Control") %>% mutate(Age = round(age)) %>% mutate(Sex = toupper(gender)) %>% rename(Participant = Subj)

# concatenate the participants. note that some of them appear to be coded as follow-ups. this is very useful, but may not be needed for the first report. they appear to be coded as such by the presence of "F1", "F2" in the participant name. convert this to a visit number (1, 2, 3, 4)
participants = full_join( m %>% select(-trial, -age, -gender) %>% distinct(), c %>% select(-trial, -age, -gender) %>% distinct() ) %>%
  mutate(`Visit Number` = ifelse(Participant %omits% "F\\d", 1, as.numeric(str_extract(str_extract(Participant, "F\\d"),"\\d"))+1)) %>% 
  mutate(`Visit Date` = date(dmy_hms(obsDate))) %>% select(-obsDate) %>% 
  mutate(`Visit ID` = Participant) %>% 
  mutate(`Visit Code` = str_extract(Participant, "F\\d")) %>% 
  mutate(Participant = str_replace(Participant, "F\\d", "")) %>% 
  mutate(Group = factor(Group, levels = c("MS","Control")))

# identify enrollment date for each participant and calculate amount of observation time
participants = participants %>% left_join(participants %>% group_by(Participant) %>% summarize(`Enrollment Date` = min(`Visit Date`, na.rm = TRUE))) %>% 
  mutate(`Time Since Enrollment (Days)` = ifelse(Group == "MS", `Visit Date` - `Enrollment Date`, 0))

# calculate amount of follow-up
followUp = participants %>% select(Group, Participant, `Time Since Enrollment (Days)`) %>% group_by(Group, Participant) %>% summarize(`Study Visits Available (n)` = n(), `Follow-Up Available (Days)` = max(`Time Since Enrollment (Days)`))

```

## Assessment protocol
Complete details of the assessment protocol are avaialble in an earlier report.[@huisinga2014] 
Briefly, subjects stood on a set of 2 custom hydraulic force platforms that translated forward together causing backward body sway and the activation of tibialis anterior (TA).
subjects stood with arms folded across the chest, eyes open, with their feet at a fixed heel-to-heel distance of 10cm.
Peak platform displacement was 4 cm, at a rate of 15 cm/s, which elicited feet-in-place balance corrections without stepping.
Surface electromyography was recorded from the right TA of all subjects with two 2.5-cm^2^ surface electrodes placed approximately 2 cm apart, with a ground electrode on the lateral condyle. 

## Data processing
Each individual EMG record was registered to the onset of platform motion and resampled.

## Exclusions
Beginning with N=TK records, TK were excluded (N=TK, MS, N=TK, Control) due to incomplete demographic information.
Characteristics of the remaining participants included in analyses are summarized in Table 1.

## Demographic and Clinical Characteristics

# Disclosures

## Funding sources and conflict of interest
This study was supported in part by National Institutes of Health grant K25HD086276.
The authors report no conflict of interest related to this study.

## Financial disclosures for the previous 12 months
Dr. McKay has received research funding or support from the National Institutes of Health. Dr. McKay has received consulting fees from BioCircuit Technologies.

# References

